---
title: "Matched cohort tangram analysis"
author: "Kyle Rove"
date: "9/19/2019"
output: html_document
---

## include libs

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
devtools::install_github("spgarbet/tangram")
library(tangram)
library(MatchIt)
library(geepack)
library(broom)
```

## Define patient population

```{r}
set.seed(1)

# patient identifier
study_id <- c(1:100)

# match on these variables
age     <- rnorm(n=length(study_id),mean=10,sd=2)                            # years
sex     <- factor(sample(c(0,1), replace=TRUE, size=length(study_id)))       # M=0, F=1
lang    <- factor(sample(c(0,1,2,3), replace=TRUE, size=length(study_id)))   # 0 = english, 1 = spanish, 2 = french, 3 = other

# outcomes
opioids <- rlnorm(n=length(study_id),0.88,0.67)                              # opioids in mg/kg
los     <- rlnorm(n=length(study_id),3.4,0.67)                               # length of stay in days
nsaid   <- factor(c(rep(0,20),rep(1,80)))                                    # 0 = no nsaid given, 1 = nsaid given
neuro   <- factor(sample(c(0,1,2,3,4), replace=TRUE, size=length(study_id))) # 0 = cervical, 1 = thoracic, 2 = lumbar, 3 = sacral, 4 = other

# comparison / arm
group   <- c(rep(0,70),rep(1,30))                                            # 0 = historical, 1 = prospective cohort

# create data frame
df     <- as_tibble(cbind(study_id,age,sex,lang,opioids,los,nsaid,neuro,group))
```

## Create matched groups, 1:1 and 2:1
# reason for two types of matching scenarios is the statistical analysis varies
# up to this point is just creation of data frames for analysis, m1.final and m2.final

```{r}
# 1:1 example
m1.out <- matchit(group ~ age + sex + lang, data = df, method = "nearest", ratio = 1)
m1.df  <- match.data(m1.out)
m1.treat <- as_tibble(df[row.names(m1.out$match.matrix),"study_id"]) %>% mutate(block = row_number())
m1.final <- as_tibble(df[m1.out$match.matrix,"study_id"]) %>% mutate(block = row_number()) %>% rbind(m1.treat) %>% left_join(m1.df,by="study_id") %>% arrange (block,group)

# 2:1 example
m2.out <- matchit(group ~ age + sex + lang, data = df, method = "nearest", ratio = 2)
m2.df  <- match.data(m2.out)
m2.treat <- as_tibble(df[row.names(m2.out$match.matrix),"study_id"]) %>% mutate(block = row_number())
m2.contr1 <- as_tibble(df[m2.out$match.matrix[,1],"study_id"]) %>% mutate(block = row_number())
m2.contr2 <- as_tibble(df[m2.out$match.matrix[,2],"study_id"]) %>% mutate(block = row_number())
m2.final <- m2.treat %>% rbind(m2.contr1,m2.contr2) %>% left_join(m2.df,by="study_id") %>% arrange (block,group) 

# change factors back to factors
m1.final$group <- factor(m1.final$group)
m1.final$sex <- factor(m1.final$sex)
m1.final$lang <- factor(m1.final$lang)
m1.final$nsaid <- factor(m1.final$nsaid)
m1.final$neuro <- factor(m1.final$neuro)
m2.final$group <- factor(m2.final$group)
m2.final$sex <- factor(m2.final$sex)
m2.final$lang <- factor(m2.final$lang)
m2.final$nsaid <- factor(m2.final$nsaid)
m2.final$neuro <- factor(m2.final$neuro)

```

## Tangram extension functions

```{r}
#
# There are several references that cover this:
# [1]	Austin PC. A critical appraisal of propensity-score matching in the medical literature between 1996 and 2003. Statist Med 2008; 27: 2037–49. doi:10.1002/sim.3150.
# [2] Fleiss JL, Levin B, Paik MC. Statistical Methods for Rates and Proportions (3rd edn). Wiley: New York, NY, 2003.
#
# 1:1 matching
#
# variables          statistical test
# -----------------  ----------------
# Numeric x Cat      paired Student's t-test [stats::t.test(x=covariate, y=arm, paired=TRUE)]
#                    Wilcoxon signed rank test [stats::wilcox.test(x=covariate, y=arm, paired=TRUE)]
#                       - preferred
#                    Cox proportional hazards models stratifying on matched groups
#                       [survival::coxph(outcome ~ covariate + strata(block), data = m1.final)]
#                       - useful for time to event analysis
# Cat X Cat          McNemar's test [stats::mcnemar.test(x=covariate, y=arm)]
#                       - this is for 2 x 2 cases only
#                       - expects factors
#                    Stuart Maxwell chi-squared test [DescTools::StuartMaxwellTest(x=covariate, y=arm)]
#                       - this is for 2 x k polytomous covariates, where k ≥ 2
#                       - expects factors
#
#
#
# many:1 matching
#
# variables          statistical test
# -----------------  ----------------
# Numeric x Cat      logistic regression with generalized estimated equations 
#                       [geepack::geeglm(formula = outcome ~ covariate, family = binomial("logit"), data = m2.final, id = block, corstr = "independence", zcor = "zcor")]
#                          - outcome must be binary numeric (not a factor)
#                          - covariate must be numeric
#                          - block must be numeric (not a factor)
#                    conditional logistic regression 
#                       [survival::clogit(outcome ~ covariate + strata(block), data = m2.final)]
# Cat x Cat          Cochran-Mantel-Haenszel chi-squared test
#                       [stats::mantelhaen.test(x=covariate, y=arm, z=block)]
#                          - strata with only 1 occurence cause errors, there should be check for this
#                          - covers 2 x 2 and 2 x >2 polytomous covariates
#

# I favor the following format:
# ==========================================================
#        N        0                  1            Statistic
#               (N=30)             (N=30)                    
# ----------------------------------------------------------
# sex:F  60    22 (73.3%)        22 (73.333%)         —       
# los    60  28.4 (20.4-27.6)  45.3 (17.4-29.8)    < 0.001    
# lang   60                                         0.630     
#    1        12 (40.0%)          8 (26.6%)                 
#    2         6 (20.0%)          7 (23.3%)                 
#    3         6 (20.0%)         10 (33.3%)                 
#    4         6 (20.0%)          5 (16.6%)                 
# nsaid  60   23 (76.7%)         30 (100.0%)        0.023     
# neuro  60                                         0.635     
#    1         7 (23.3%)         10 (33.3%)                 
#    2         7 (23.3%)          6 (20.0%)                 
#    3         2 (6.7%)           5 (16.6%)                 
#    4         9 (30.0%)          6 (20.0%)                 
#    5         5 (16.7%)          3 (10.0%)                 
# ==========================================================

summarize_numerical <- function(table,
                                row,                     # numerical covariate
                                column,                  # dichotomous outcome
                                block,                   # vector of block assignments 
                                pref_test="wilcox.test", # preferred statistical test
                                cell_style,
                                pformat=NULL,
                                collapse_single=TRUE,
                                test=TRUE,
                                ...)
{

  # get data
  covariate <- row$data
  outcome <- column$data
  n_matched <- length(block) / length(levels(as.categorical(block)))
  
  # make the df and sort it
  df <- data.frame(covariate=as.numeric(covariate), outcome=as.numeric(levels(factor(outcome,levels=c(0,1))))[outcome], block=as.numeric(block) )
  df <- df %>% arrange(block,outcome)
    
  # first branch point is whether data is matched 1:1 or 1:many
  if (n_matched == 2) {

    # paired Student's t-test
    if (pref_test == "t.test") {
      # run test
      stat <- NA
      stat <- t.test(x=df$covariate[df$outcome == 0], y=df$covariate[df$outcome == 1], paired=TRUE)
    
      # get p value
      if (length(stat) > 1) {
        p_val <- cell_style[['p']](p = broom::tidy(stat)$p.value)
      } else {
        p_val <- cell_style[['p']](p = NA)
      }
      
    # Wilcoxon signed rank test
    } else if (pref_test == "wilcox.test") {
      # run test
      stat <- NA
      stat <- wilcox.test(x=df$covariate[df$outcome == 0], y=df$covariate[df$outcome == 1], paired=TRUE)

      # get p value
      if (length(stat) > 1) {
        p_val <- cell_style[['p']](p = broom::tidy(stat)$p.value)
      } else {
        p_val <- cell_style[['p']](p = NA)
      }
      
    # Cox proportional hazards model stratefied on matched pairs
    } else if (pref_test == "coxph") {
      # run regression
      stat <- NA
      stat <- survival::coxph(outcome ~ covariate + strata(block), data = df)
    
      # get p value
      if (length(stat) > 1) {
        p_val <- cell_style[['p']](p = broom::tidy(stat)$p.value)
      } else {
        p_val <- cell_style[['p']](p = NA)
      }
      
    # any other specified preferred_test is invalid
    } else {
      stat  <- NA
      p_val <- cell_style[['p']](p = NA)
    }
    
  } else if (n_matched > 2) {
    
    # logistic regression with generalized estimating equations
    if (pref_test == "geeglm") {
      # run regression
      stat <- NA
      stat <- suppressWarnings(geepack::geeglm(formula = outcome ~ covariate, family = binomial("logit"), data = df, id = block, corstr = "independence", zcor = "zcor"))
    
      # get p value
      if (length(stat) > 1) {
        p_val <- cell_style[['p']](p = broom::tidy(stat)$p.value)
      } else {
        p_val <- cell_style[['p']](p = NA)
      }
      
    # conditional logistic regression   
    } else if (pref_test == "clogit") {
      # run regression
      stat <- NA
      stat <- survival::clogit(outcome ~ covariate + strata(block), data = df)
    
      # get p value
      if (length(stat) > 1) {
        p_val <- cell_style[['p']](p = broom::tidy(stat)$p.value)
      } else {
        p_val <- cell_style[['p']](p = NA)
      }
    
    # any other specified preferred_test is invalid
    } else {
      stat  <- NA
      p_val <- cell_style[['p']](p = NA)
    }
  }
  
  # Get column names
  col_categories <- levels(outcome)
  
  # Get n values
  n_labels <- lapply(col_categories, FUN=function(cat_name){
    x <- df$covariate[df$outcome == cat_name]
    cell_n(length(unique(x)), subcol=cat_name)
  })
  
  # Build table
  table                                                %>%
    # Create Headers
    row_header(derive_label(row))                      %>%
    col_header("N", col_categories, "Test Statistic")  %>%
    col_header("",  n_labels,       ""              )  %>%
    # Add the First column of summary data as an N value
    add_col(cell_n(length(unique(outcome))))           %>%
    # Now add median (IQR) for the counts
    table_apply(col_categories, FUN=
                  function(tbl, cat_name) {
                    # Compute each data set
                    x  <- df$covariate[df$outcome == cat_name]
                    
                    # Add a column that is a quantile
                    add_col(tbl, cell_style[['iqr']](x, row$format, na.rm=TRUE))
                  })                                   %>%
    # Now add a statistical test for the final column
    add_row(p_val)

  table
}

summarize_categorical <- function(table,
                                  row,                  # categorical covariate
                                  column,               # dichotomous outcome
                                  block,                # vector of block assignments 
                                  cell_style,
                                  pformat=NULL,
                                  collapse_single=TRUE,
                                  test=TRUE,
                                  ...)
{
  
  # get data
  covariate <- as.categorical(row$data)
  outcome <- as.categorical(column$data)
  block <- as.categorical(block)
  n_matched <- length(levels(as.categorical(block)))

  # make the df and sort it
  df <- data.frame(var=as.numeric(covariate), outcome=as.numeric(levels(factor(outcome,levels=c(0,1))))[outcome], block=as.numeric(block) )
  df <- df %>% arrange(block,outcome)
  
  # define tables and validate
  grid          <- table(covariate, outcome, block, useNA="no")
  grid2         <- table(covariate, outcome, useNA="no")
  validrow      <- which(!apply(grid,1,FUN = function(x){all(x == 0)}))
  validcol      <- which(!apply(grid,2,FUN = function(x){all(x == 0)}))
  validblocks   <- which(!apply(grid,3,FUN = function(x){all(x == 0)}))
  invalidstatum <- which(apply(grid,1,FUN = function(x){sum(ifelse(x==1,1,0))})==1)
  nrow          <- dim(grid)[1]
  ncol          <- dim(grid)[2]
  ntuple        <- dim(grid)[3]
  denominators  <- matrix(rep(colSums(grid2), nrow), ncol=ncol, byrow=TRUE)
  
  # first branch point is whether data is matched 1:1 or 1:many and how many levels the categorical variable has
  if (n_matched == 2 && length(levels(covariate)) == 2) {

    # McNemar's test
    stat <- NA
    stat <- mcnemar.test(x=df$covariate[df$outcome == 0], y=df$covariate[df$outcome == 1])
    
    # get p value
    if (length(stat) > 1) {
      p_val <- cell_style[['p']](p = broom::tidy(stat)$p.value)
    } else {
      p_val <- cell_style[['p']](p = NA)
    }
    
  } else if (n_matched == 2 && length(levels(covariate)) > 2) {
    
    # Stuart Maxwell chi-squared test
    stat <- NA
    stat <- StuartMaxwellTest(x=df$covariate[df$outcome == 0], y=df$covariate[df$outcome == 1])
    
    # get p value
    if (length(stat) > 1) {
      p_val <- cell_style[['p']](p = broom::tidy(stat)$p.value[[1]])
    } else {
      p_val <- cell_style[['p']](p = NA)
    }
    
  } else if (n_matched > 2) {
    
    # Cochran-Mantel-Haenszel chi-squared test
    stat <- NA
    stat <- if(length(validrow) < 2 || length(validcol) < 2 || length(validblocks) < 1 || length(invalidstatum) > 0) NA else mantelhaen.test(covariate,outcome,block)

    # get p value
    if (length(stat) > 1) {
      p_val <- cell_style[['p']](p = broom::tidy(stat)$p.value)
    } else {
      p_val <- cell_style[['p']](p = NA)
    }
    
  }
  
  # start collecting row labels
  rowlabels     <- rownames(grid2)
  
  # Compute overall N values for each category
  subN <- lapply(colnames(grid2), FUN=function(cat)
    cell_style[['n']](sum(column$data == cat, na.rm=TRUE), subcol=cat, possible=length(column$data), hdr=TRUE, ...)
  )
  
  # Collapse to a single line when requested for 2 binomial factors
  if(collapse_single && dim(grid2)[1]<=2)
  {
    # Why is this so difficult?
    
    # More complex name derivation
    name <- row$name()
    try({
      l2 <- attr(row$data, "label")
      if(!is.null(l2)) {name<-l2}
    })
    
    pos <- dim(grid2)[1]
    
    # Select part of grid table, then do all the munging to get it back in form
    x <- matrix(grid2[pos,], nrow=1)
    colnames(x) <- colnames(grid2)
    rownames(x) <- name
    grid2 <- x
    denominators <- matrix(denominators[pos,], nrow=1)
    nrow <- 1
  }
  else # Give a good indent otherwise
  {
    rownames(grid2)   <- lapply(rownames(grid2), FUN=function(x) paste("  ", x))
  }
  
  # Build the table
  # Column Headers
  if(test) {
    table <- col_header(table, "N", colnames(grid2), "Test Statistic")
    table <- col_header(table, "", subN, "")
  } else {
    table <- col_header(table, "N", colnames(grid2))
    table <- col_header(table, "", subN)
  }
  
  # Row Headers
  if(nrow > 1) table <- row_header(table, derive_label(row)) # Deal with single
  for(nm in rownames(grid2)) table <- row_header(table, nm)
  
  # The N value
  table <- add_col(table, cell_style[['n']](sum(!is.na(row$data)), possible=length(row$data), ...))

  # Now loop the grid2 into the table as a fraction
  for(j in 1:ncol)
  {
    if(nrow > 1) table <- add_row(table, "")
    format <- if(is.na(row$format) || is.null(row$format)) format_guess(as.vector(grid2/denominators)) else row$format
    for(i in 1:nrow)
    {
      if(denominators[i,j] == 0) {
        table <- add_row(table, "")
      } else {
        table <- add_row(table,
                   cell_style[['fraction']](
                     grid2[i,j], denominators[i,j],
                     format=format,
                     subcol=colnames(grid2)[i],
                     subrow=rownames(grid2)[j]))
      }
    }
    table <- new_col(table)
  }
  
  # add the stats
  table <- add_row(table,p_val)
    
  # add blanks in test column if sub-categories
  if(nrow > 1) table <- add_row(table, rep("", nrow))

  table
}

psm <- list(
  Type        = hmisc_data_type,
  Numerical   = list(
    Categorical = summarize_numerical 
  ),
  Categorical = list(
    Categorical = summarize_categorical
  ),
  Cell        = hmisc_cell,
  Footnote    = "Numerical summary is median (IQR). Categorical is N (%). ^1^Logistic regression with GEE. ^2^Cochran Mantel Maenszel."
)

psm[['Cell']][['fraction']] <- function(numerator, denominator, format=3, ...) {
  paste0(numerator,' (',render_f(100*numerator/denominator, format),'%)')
}

psm[['Cell']][['iqr']] <- function(x,format,na.rm,names=TRUE,type=8,msd, quant = c(0.25, 0.75), ...) {
  iqr <- quantile(x, c(0.25, 0.75), na.rm, names, type)
  m <- median(x, na.rm)
  if(is.na(format)) format <- format_guess(y)
  iqr_formatted <- sapply(iqr, function(x) render_f(x, format))
  paste0(render_f(m, format),' (',paste0(iqr_formatted,collapse="–"),')')
}

psm[['Cell']][['p']] <-  function(p, pformat="%1.3f", include_p=FALSE) {
  if(class(pformat) == "function") pformat(p)
  
  if(is.na(p) || is.nan(p) || p <0 || p>1) return("—")
  
  y <- render_f(p, pformat)
  
  # Check for all zeros once formated
  test <- grep("[^0\\.]+", y)
  if(length(test) > 0) {
    if(include_p) paste0("p = ",y) else y
  }
  else {
    if(include_p) paste0("p < ", substr(y, 1, nchar(y)-1), "1") else paste0("<", substr(y, 1, nchar(y)-1), "1")
  }
}
```

## Tangram calls, analysis with propensity matched historical controls and prospective treated patients

```{r}
# 1:1 example
table1 <- tangram ( group              # 2-level categories are typically only dependent variable to my knowledge in propensity score matching analyses
                          ~ age[1]     # numeric
                          + sex        # binary categorical
                          + lang       # multi-level categorical
                          + opioids[1] # numeric
                          + los[1]     # numeric
                          + nsaid      # binary categorical
                          + neuro,     # multi-level categorical
                          data = m1.final, id = "table1", pref_test = "wilcox.test", block = m1.final$block, test = TRUE, transform = psm )
                              # psm = propensity score matching, seems like a good name for this analysis
                              # need to pass block vector (block = X) to the transform functions

# 2:1 example
table2 <- tangram ( group              # 2-level categories are typically only dependent variable to my knowledge in propensity score matching analyses
                          ~ age[1]     # numeric
                          + sex        # binary categorical
                          + lang       # multi-level categorical
                          + opioids[1] # numeric
                          + los[1]     # numeric
                          + nsaid      # binary categorical
                          + neuro,     # multi-level categorical
                          data = m2.final, id = "table1", pref_test = "glmgee", block = m2.final$block, test = TRUE, transform = psm )
                              # psm = propensity score matching, seems like a good name for this analysis
                              # need to pass block vector (block = X) to the transform functions
```